Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-26 paper/paper.tex
Fix typos and change language name to be more descriptive

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,screen=true,
\ifpublic review=false\else,review=true\fi
  ,anonymous=\ifanonymous true\else false\fi]{acmart}
\usepackage{fontspec}
\newfontfamily\boxedsymbols{DejaVu Sans}
\usepackage{ottalt}
\usepackage{minted}
\usepackage{xspace}
\usepackage{tcolorbox}
\usepackage[para]{footmisc}
\definecolor{lightgray}{gray}{0.85}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^H$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relation is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed or polymorphic lambda calculus, a streamlined
presentation of proof by logical relation for a dependently language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a full universe
hierarchy, booleans, and an intensional identity type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

@@

Our goal with this work is to demonstrate how the proof technique of
\emph{syntatic logical relations} can be applied to reason about dependent
type theories.  Logical relations are a powerful proof technique for reasoning
about lambda calculi and their operational semantics. They have been used to
show diverse properties such as strong normalization, contextual equivalence,
representation independence, noninterference, compiler correctness, and the
decidability of conversion algorithms. \scw{add citations? emphasize that we are working
with syntactic logical relations.}

Tutorial material on logical relations~\citep{skorstengaard2019introduction,
  harpertait, harperkripke, pierce2002types,
  pierce2004advanced,harper2016practical} is primarily focussed on systems
with simple or polymorphic types. In that context, syntactic logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation. However, neither of these techniques can be used to define a
logical relation in the context of a predicative dependent type theory, so a
young researcher might be excused for thinking that proofs that use logical
relations are not applicable for such languages.

@@

We have streamlined our proof through a number of means: the careful selection
of the features that we include in the object type type system and the results
that we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, but we do include propositionality
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hiercharchy and
type-level computation, to demonstrate the logical strength of the
approach. % \scw{Say that we don't have impredicativity, but neither does anyone
  % else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

The result of our work is a development that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set theory notation and terminology so that
the material is accessible to readers with general mathematical background,
That said, our explanations do not stray too far away from our proof scripts.
We label each lemma directly to their counterpart in the proof script,
anticipating that readers may wish to reference and validate our formalism.
Our typeset representation are purposefully made to directly follow our
mechanical proofs, while avoiding, as much as possible, the introduction
of artifacts that are specific to theorem provers.

Not only does this connection aid readers that wish to, like us, adopt proof
assistants for their day-to-day use. We also find that this precision is
important for conveying the proof techique itself. Unlike properties that are
derivable through syntactic means, proofs by logical relation make demands on
the strength of the logic in which they are expressed. An informal proof that
attempts to be agnostic or ambiguous about the underlying
metatheory % may struggle to even
% convey the validity of their definition and
requires substantial
effort from the reader to even understand whether it is theoretically
possible to encode such proofs in a given ambient logic, such as those found
in modern proof assistants.

@@

In Section~\ref{sec:extension}, extends the definition of the logical relation
to include terms with free variables, and proves that every well-typed term
has a $\beta$-normal form. After laying the foundation for handling type-level
computation in Section~\ref{sec:logreldep}, this extension requries mechanical
changes that directly mirror an analogous extension for a simply typed
language. We use this section to demonstrate the idea that once we have
established the base techniques, we can port ideas that have been studied in
the context of a simpler system to the dependently-typed setting.  We believe
it is possible to extend our logical relation to a Kripke-style and relational
model in a similar fashion.

In Section~\ref{sec:eta} we show how to incorporate the function $\eta$-law
into the conversion rule for \lang{}.

@@

@@

% To derive consistency, it suffices to restrict $[[S]]$ to the set of
% terms that reduce to closed terms.
\Rref{I-Void, I-Bool} capture terms that \emph{behave like} the inhabitants
of the $[[Void]]$ and $[[Bool]]$ types under an empty context. For
example, the $[[Void]]$ type should not have any inhabitants under the
empty context, where as the $[[Bool]]$ type corresponds to terms that
evaluate to $[[true]]$ or $[[false]]$. However, the characterization
of $[[Bool]]$ in our logical relation does not always correspond to
well-typed or even closed terms. For example, we have $[[if false var 0
true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
  true \/ a =>+ false }]]$ since it evalutes to $[[true]]$.
The independence of syntactic typing in our logical relation allows
our semantic typing definition in Section~\ref{sec:logrelproof} to be
meaningful on its own. Furthermore, not having to embed scoping
information into the logical relation avoids extra bookkeeping and the
need for a Kripke-style logical relation.
\yl{Not sure what to cite from Derek Dreyer. I know his blog post
  about semantic type soundness but is there a good paper to cite? one
of the rust papers?}

% unsurprising; when considering only closed terms, the empty type
% should not be inhabited and therefore corresponds to the empty set,
% whereas the boolean type is semantically inhabited by terms that
% evaluate to the boolean values $[[true]]$ or $[[false]]$.
\Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance. \Rref{I-Red} enables us to reduce our types in order
to assign meanings. Recall the type expression $[[(\ var 0)
Bool]]$. \Rref{I-Red} says to know that
$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Red}
and \rref{I-Bool}.

\Rref{I-Pi} is the most complicated rule in our logical
relation. Instead of explaining it directly, we consider the following
simplified admissible \rref{I-PiAlt} that follows directly from \rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}[I-PiAlt admissibility]
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
, S0 ) | a in S , Interp I i B { a } S0 }]]$.
\end{proof}
Compared to the \rref{I-Pi}, which contains the mysterious relation
$[[R]]$, \rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The precondition of
\rref{I-PiAlt} says that
the function type $[[Pi A B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation. In its conclusion, we say
that a term $[[b]]$ belongs to an interpretation of $[[Pi A B]]$ if
for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
application form $[[b a]]$ belongs to all possible interpretations of
$[[B {a}]]$ (the pre-condition ensures at least one interpretation
exists for each $[[B {a}]]$ where $[[a in S]]$).

@@

  Intuitively, a substitution $[[rho]]$
  that closes over a context should always induce a truncated
  substituion for each truncated context. As a result, if
  some term $[[b]]$ is semantically well-typed under some truncated
  context $[[G]]$, then it must also be semantically well-typed under
  the context $[[G]]$ up to shifting.
%
  Knowing that $[[G |= G n {up Suc n} : Set i]]$ holds, we can then
  apply Lemma~\ref{lemma:setinv} to show that
  $[[InterpR i G n < up Suc n > { rho } S]]$. It now
  suffices to show $[[rho n in S]]$, but that is immediate from the
  definition of $[[rho |= G]]$.
\end{proof}

@@

Furthermore, if we know that an application of a term to a variable
has a normal form, then we know tha the term must have a normal form.
\begin{lemma}[Wn extensionality\footnote{\dotv{normalform.v}{ext\_wn}}]
  \label{lemma:extwn}
  If $[[wn (a var i)]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn (a
  var i)]]$. The conclusion follows from Lemmas~\ref{lemma:parantirenaming} and
  \ref{lemma:parnenf}.
\end{proof}

@@

In \citet{nbeincoq,decagda,martin-lof-a-la-coq}, a relational model is
needed to justify the $\eta$-law for functions. Instead, we prove
confluence for our parallel reduction extended with $\eta$-law for functions
and continue using a logical predicate (i.e. a 1-ary logical relation) to justify
$\eta$-law as part of our equational theory.
A well-known issue with our approach is the failure of syntactic
confluence when the lambda term contains type annotations. A simple
counterexample is $[[\- B ((\- A a) {up 1} var 0)]]$; depending on
whether \rref{P-AbsEta} is performed on the whole term or
\rref{P-AppAbs} is used on the inner $\beta$ redex, we end up with the
terms $[[\- B a]]$ or $[[\- A a]]$, where $[[A]]$ and $[[B]]$ are not
necessarily syntactically equal terms. \citet{choudhury:ddc} resolve
this problem by stating their confluence result in terms of an
equivalence relation that quotients out parts of the terms that are
computationally irrelevant; the annotations of lambda terms are
ignored since the behavior of a lambda term is not affected by its
type annotation. We believe the same approach is applicable to our
proof if we add type annotations to our lambda terms.

@@

Finally, in set theory, to show that two sets $[[S0]]$ and $[[S1]]$
are equal, it suffices to show the extensional property that $\forall
x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
occassionally in our presented proofs.
In Coq, sets of terms ($[[PowerSet STm]]$) is encoded as the type \texttt{tm ->
Prop}, a predicate over \lang{} terms.
In axiom-free Coq, predicates do not come with the extensionality
property. Given two predicates $P$ and $Q$, we can not conclude that
$P = Q$ when given a proof of $\forall
x, P(x) \iff Q(x)$. While
predicate extensionality is not necessary in our development, we
assume it as an axiom since it is known to be consistent with Coq's
metatheory and helps bridge the gap between our mechanization and the
proofs we present.

@@

We use the Autosubst 2 framework to specify our syntax in HOAS and
produce Coq syntax files in de Bruijn representation. Additionallly,
Autosubst 2 provides a powerful tactic \texttt{asimpl} that
can be used to prove the equivalence of two terms constructed using
the primitive operators provided by the framework. This greatly
simplifies the reasoning about substitution in our development as
almost all substitution related properties about the syntax are
immediately discharged by \texttt{asimpl}.

@@

The automation provided by CoqHammer not only gives us a proof that is shorter
and more resiliant to changes, but also gives useful documentation for readers
who wish to understand the mechanized proof. Although automation performs
extensive search, it does not use lemmas or invert derivations that are not
specified in the \texttt{use} or \texttt{inv} flag.

@@

More recent works such as \citet{Abel12}
and\citet{abel2008betaeta} make their definitions more explicit and
precise and thus easier to encode in proof assistants. Our definition
of logical relation is reminiscent of their definition of a
semantic universe hierarchy, though our logical relation is defined to
be closed under expansion with respect to parallel reduction rather
than weak head reduction. Furthermore, \citet{Abel12} and
\citet{abel2008betaeta} use the semantic universe hierachy as a
measure to define Kripke-style logical relations, from which they
derive the correctness of their conversion algorithms. In our work, we
use the semantic universe hierarchy directly in our definition of
semantic typing because it is sufficient
for deriving consistency and later normalization.

@@

\citet{decagda} mechanizes in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$-law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facillitate the reuse of their definition. The
logical relation is defined using the induction-recursion scheme,
which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} manages to encode the logical relation
from \citet{decagda} in the predicative fragment of Coq and further
extends the decidability of type conversion result
from~\citet{decagda} to the decidability type checking of a
bidrectional type system.

>>>>

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,screen=true,
\ifpublic review=false\else,review=true\fi
  ,anonymous=\ifanonymous true\else false\fi]{acmart}
\usepackage{fontspec}
\newfontfamily\boxedsymbols{DejaVu Sans}
\usepackage{ottalt}
\usepackage{minted}
\usepackage{xspace}
\usepackage{tcolorbox}
\usepackage[para]{footmisc}
\definecolor{lightgray}{gray}{0.85}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^{U\mathbb{B}\ottkw{Id}}$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relation is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed or polymorphic lambda calculus, a streamlined
presentation of proof by logical relation for a dependently typed language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a full universe
hierarchy, booleans, and an intensional identity type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

@@

Our goal with this work is to demonstrate how the proof technique of
\emph{syntactic logical relations} can be applied to reason about dependent
type theories.  Logical relations are a powerful proof technique for reasoning
about lambda calculi and their operational semantics. They have been used to
show diverse properties such as strong normalization, contextual equivalence,
representation independence, noninterference, compiler correctness, and the
decidability of conversion algorithms. \scw{add citations? emphasize that we are working
with syntactic logical relations.}

Tutorial material on logical relations~\citep{skorstengaard2019introduction,
  harpertait, harperkripke, pierce2002types,
  pierce2004advanced,harper2016practical} is primarily focused on systems
with simple or polymorphic types. In that context, syntactic logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation. However, neither of these techniques can be used to define a
logical relation in the context of a predicative dependent type theory, so a
young researcher might be excused for thinking that proofs that use logical
relations are not applicable for such languages.

@@

We have streamlined our proof through a number of means: the careful selection
of the features that we include in the object type system and the results
that we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, but we do include propositional
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hierarchy and
type-level computation, to demonstrate the logical strength of the
approach. % \scw{Say that we don't have impredicativity, but neither does anyone
  % else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

The result of our work is a development that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set theory notation and terminology so that
the material is accessible to readers with general mathematical background,
That said, our explanations do not stray too far away from our proof scripts.
We label each lemma directly to its counterpart in the proof script,
anticipating that readers may wish to reference and validate our formalism.
Our typeset representations are purposefully made to directly follow our
mechanical proofs, while avoiding, as much as possible, the introduction
of artifacts that are specific to theorem provers.

Not only does this connection aid readers that wish to, like us, adopt proof
assistants for their day-to-day use. We also find that this precision is
important for conveying the proof technique itself. Unlike properties that are
derivable through syntactic means, proofs by logical relation make demands on
the strength of the logic in which they are expressed. An informal proof that
attempts to be agnostic or ambiguous about the underlying
metatheory % may struggle to even
% convey the validity of their definition and
requires substantial
effort from the reader to even understand whether it is theoretically
possible to encode such proofs in a given ambient logic, such as those found
in modern proof assistants.

@@

In Section~\ref{sec:extension}, extends the definition of the logical relation
to include terms with free variables, and proves that every well-typed term
has a $\beta$-normal form. After laying the foundation for handling type-level
computation in Section~\ref{sec:logreldep}, this extension requires mechanical
changes that directly mirror an analogous extension for a simply typed
language. We use this section to demonstrate the idea that once we have
established the base techniques, we can port ideas that have been studied in
the context of a simpler system to the dependently-typed setting.  We believe
it is possible to extend our logical relation to a Kripke-style and relational
model in a similar fashion.

In Section~\ref{sec:betaeta} we show how to incorporate the function $\eta$-law
into the conversion rule for \lang{}.

@@

In Section~\ref{sec:discuss}, we explain the various factors that contribute
to our short proof.

@@

% To derive consistency, it suffices to restrict $[[S]]$ to the set of
% terms that reduce to closed terms.
\Rref{I-Void, I-Bool} capture terms that \emph{behave like} the inhabitants
of the $[[Void]]$ and $[[Bool]]$ types under an empty context. For
example, the $[[Void]]$ type should not have any inhabitants under the
empty context, where as the $[[Bool]]$ type corresponds to terms that
evaluate to $[[true]]$ or $[[false]]$. However, the characterization
of $[[Bool]]$ in our logical relation does not always correspond to
well-typed or even closed terms. For example, we have $[[if false var 0
true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
  true \/ a =>+ false }]]$ since it evaluates to $[[true]]$.
The independence of syntactic typing in our logical relation allows
our semantic typing definition in Section~\ref{sec:logrelproof} to be
meaningful on its own. Furthermore, not having to embed scoping
information into the logical relation avoids extra bookkeeping and the
need for a Kripke-style logical relation.
\yl{Not sure what to cite from Derek Dreyer. I know his blog post
  about semantic type soundness but is there a good paper to cite? one
of the rust papers?}

% unsurprising; when considering only closed terms, the empty type
% should not be inhabited and therefore corresponds to the empty set,
% whereas the boolean type is semantically inhabited by terms that
% evaluate to the boolean values $[[true]]$ or $[[false]]$.
\Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresponds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance. \Rref{I-Red} enables us to reduce our types in order
to assign meanings. Recall the type expression $[[(\ var 0)
Bool]]$. \Rref{I-Red} says to know that
$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Red}
and \rref{I-Bool}.

\Rref{I-Pi} is the most complicated rule in our logical
relation. Instead of explaining it directly, we consider the following
simplified admissible \rref{I-PiAlt} that follows directly from \rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}[I-PiAlt admissibility]
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
, S0 ) | a in S , Interp I i B { a } S0 }]]$.
\end{proof}
Compared to the \rref{I-Pi}, which contains the mysterious relation
$[[R]]$, \rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The precondition of
\rref{I-PiAlt} says that
the function type $[[Pi A B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituting $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation. In its conclusion, we say
that a term $[[b]]$ belongs to an interpretation of $[[Pi A B]]$ if
for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
application form $[[b a]]$ belongs to all possible interpretations of
$[[B {a}]]$ (the pre-condition ensures at least one interpretation
exists for each $[[B {a}]]$ where $[[a in S]]$).

@@

  Intuitively, a substitution $[[rho]]$
  that closes over a context should always induce a truncated
  substitution for each truncated context. As a result, if
  some term $[[b]]$ is semantically well-typed under some truncated
  context $[[G]]$, then it must also be semantically well-typed under
  the context $[[G]]$ up to shifting.
%
  Knowing that $[[G |= G n {up Suc n} : Set i]]$ holds, we can then
  apply Lemma~\ref{lemma:setinv} to show that
  $[[InterpR i G n < up Suc n > { rho } S]]$. It now
  suffices to show $[[rho n in S]]$, but that is immediate from the
  definition of $[[rho |= G]]$.
\end{proof}

@@

Furthermore, if we know that an application of a term to a variable
has a normal form, then we know the term must have a normal form.
\begin{lemma}[Wn extensionality\footnote{\dotv{normalform.v}{ext\_wn}}]
  \label{lemma:extwn}
  If $[[wn (a var i)]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn (a
  var i)]]$. The conclusion follows from Lemmas~\ref{lemma:parantirenaming} and
  \ref{lemma:parnenf}.
\end{proof}

@@

In \citet{nbeincoq,decagda,martin-lof-a-la-coq}, a relational model is
needed to justify the $\eta$-law for functions. Instead, we prove
confluence for our parallel reduction extended with $\eta$-law for functions
and continue using a logical predicate (i.e. a unary logical relation) to justify
$\eta$-law as part of our equational theory.
A well-known issue with our approach is the failure of syntactic
confluence when the lambda term contains type annotations. A simple
counterexample is $[[\- B ((\- A a) {up 1} var 0)]]$; depending on
whether \rref{P-AbsEta} is performed on the whole term or
\rref{P-AppAbs} is used on the inner $\beta$ redex, we end up with the
terms $[[\- B a]]$ or $[[\- A a]]$, where $[[A]]$ and $[[B]]$ are not
necessarily syntactically equal terms. \citet{choudhury:ddc} resolve
this problem by stating their confluence result in terms of an
equivalence relation that quotients out parts of the terms that are
computationally irrelevant; the annotations of lambda terms are
ignored since the behavior of a lambda term is not affected by its
type annotation. We believe the same approach is applicable to our
proof if we add type annotations to our lambda terms.

@@

Finally, in set theory, to show that two sets $[[S0]]$ and $[[S1]]$
are equal, it suffices to show the extensional property that $\forall
x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
occasionally in our presented proofs.
In Coq, sets of terms ($[[PowerSet STm]]$) is encoded as the type \texttt{tm ->
Prop}, a predicate over \lang{} terms.
In axiom-free Coq, predicates do not come with the extensionality
property. Given two predicates $P$ and $Q$, we can not conclude that
$P = Q$ when given a proof of $\forall
x, P(x) \iff Q(x)$. While
predicate extensionality is not necessary in our development, we
assume it as an axiom since it is known to be consistent with Coq's
metatheory and helps bridge the gap between our mechanization and the
proofs we present.

@@

We use the Autosubst 2 framework to specify our syntax in HOAS and
produce Coq syntax files in de Bruijn representation. Additionally,
Autosubst 2 provides a powerful tactic \texttt{asimpl} that
can be used to prove the equivalence of two terms constructed using
the primitive operators provided by the framework. This greatly
simplifies the reasoning about substitution in our development as
almost all substitution related properties about the syntax are
immediately discharged by \texttt{asimpl}.

@@

The automation provided by CoqHammer not only gives us a proof that is shorter
and more resilient to changes, but also gives useful documentation for readers
who wish to understand the mechanized proof. Although automation performs
extensive search, it does not use lemmas or invert derivations that are not
specified in the \texttt{use} or \texttt{inv} flag.

@@

More recent works such as \citet{Abel12}
and\citet{abel2008betaeta} make their definitions more explicit and
precise and thus easier to encode in proof assistants. Our definition
of logical relation is reminiscent of their definition of a
semantic universe hierarchy, though our logical relation is defined to
be closed under expansion with respect to parallel reduction rather
than weak head reduction. Furthermore, \citet{Abel12} and
\citet{abel2008betaeta} use the semantic universe hierarchy as a
measure to define Kripke-style logical relations, from which they
derive the correctness of their conversion algorithms. In our work, we
use the semantic universe hierarchy directly in our definition of
semantic typing because it is sufficient
for deriving consistency and later normalization.

@@

\citet{decagda} mechanizes in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$-law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facilitate the reuse of their definition. The
logical relation is defined using the induction-recursion scheme,
which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} manages to encode the logical relation
from \citet{decagda} in the predicative fragment of Coq and further
extends the decidability of type conversion result
from~\citet{decagda} to the decidability type checking of a
bidirectional type system.
